How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2023-12-29 00-author.Rmd
Add files via upload
#general update of all files
Add files via upload
# with new chapter on planning

John Measey is Associate Professor of Biological Sciences at Stellenbosch University. He has authored or co-authored more than 200 peer reviewed scientific papers and book chapters, and five books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals. He has graduated more than 20 postgraduate students, and his blog on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the beautiful Western Cape, South Africa.

@@

>>>>

John Measey is Professor at the Institute of Biodiversity, Yunnan University, and Professor Extraordinary at Stellenbosch University. He has authored or co-authored [more than 250](https://orcid.org/0000-0001-9939-7615) peer reviewed scientific papers and book chapters, and six books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for three other journals. He has graduated [more than 25 postgraduate students](http://john.measey.com/About-Me/Measey-Mini-CV), and [his blog](john.measey.com/blog) on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the Spring City, Kunming in Yunnan Province, China and in Cape Town, South Africa.

@@

## How to cite {-}
Measey, J. (2022) *How to write a PhD in Biological Sciences: a guide for the uninitiated*. CRC Press, Boca Raton. Online Version: **`r Sys.Date()`**  ISBN: 9781032080208

### 2023-12-29 1.01-beforeyoustart.Rmd
Add files via upload
# with new chapter on planning

## Think about your career before you start your PhD
The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do. Although this does sound ludicrous, there are many people in academia who did PhDs for this reason, or simply because they were offered an opportunity to do a PhD and felt that they couldn’t turn it down.

@@

## Don't rely on becoming an academic
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full-time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well-funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs, with those tenured academics earning good, secure salaries at the top of the pyramid, while increasing numbers of insecure poorly paid doctoral students and post-docs do most of the actual work [@afonso2014academia].

@@

>>>>

## Think about your career before you start your PhD {#career}
The last thing a PhD should be is a stop gap for you to fill your time because you can’t think of anything else to do. Although this does sound ludicrous, there are many people in academia who did PhDs for this reason, or simply because they were offered an opportunity to do a PhD and felt that they couldn’t turn it down.

@@

## Don't rely on becoming an academic {#relyacademic}
Academia is getting more competitive. Whereas once it was almost certain that you would be employed as an academic after finishing your PhD, this is no longer the case. Indeed, the number of people who get PhDs is greatly increasing the world over, and most are in direct competition for a decreasing amount of academic posts. In 2019, 39% of respondents to a survey said that the most important reason they decided to enrol in a PhD programme was to pursue an academic career [@woolston2019phds]. Further, 56% wanted academia as their first choice for a career. Yet in Belgium, for example, the numbers of PhD graduates entering the market nearly doubled (in the first 15 years of the 21st century) while academic positions remained static [@levecque2017work]. Universities appear to be exploiting this excess of qualified people to hire them on contracts, reducing the number of full-time positions. Essentially, this means that in most places your time as a PhD student, postdoc and junior lecturer will keep you outside the employers core staff. Joining those on the inside (with well-funded posts) becomes increasingly more difficult. Some have suggested that the world of academia is becoming more like a gang producing drugs, with those tenured academics earning good, secure salaries at the top of the pyramid, while increasing numbers of insecure poorly paid doctoral students and post-docs do most of the actual work [@afonso2014academia].

@@

>I don't think students are made to understand how hard it is to do research. And how very, very hard it is to do important research. It's a lot harder than taking even very demanding courses. What makes it difficult is that research is immersion in the unknown. We just don't know what we're doing. We can't be sure whether we're asking the right question or doing the right experiment until we get the answer or the result. 
>
>Martin A. Schwartz [-@schwartz2008importance]

## The importance of being productively stupid

There will be many points during your PhD when you feel stupid because you don't know what is the right thing to do. This is normal. Indeed, when you feel that you do know exactly what you are doing, this might be a more worrying moment that gives you more pause for thought. You will need to get your head around consistently feeling stupid and feel comfortable with saying 'I don't know' without it causing you to feel like giving up. Schwartz [-@schwartz2008importance] talks about the importance of embracing that feeling of absolute stupidity when faced with many aspects of science. Facing that stupidity on a daily basis is necessary in order to have the correct attitude to your own studies. You can only really expect that your thesis will make you a little less stupid in the subject area of your choice. However, it may end up with you embracing more stupidity in ever increasing areas. When you feel comfortable with this stupidity, you'll know that you have achieved one of the most important qualities of a scientist.
